Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-10 analysis/paper/paper.Rmd
work on discussion

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, particularly trade ornaments, during the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. There are significant differences especially between Chinese contact and either pre-European or post-European (P-value < 0.01). We also observe no obvious changes in the clay paste across different phases in relation to the increasing standardization of the pottery shape. The more standardized shape in the Chinese period might indicate the appearance of specialized group. However, when we look at the coef

The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases. To figure out whether this change is related to the function of pot, we also attempt to conduct a preliminary lipid analysis for potsherds. However, we did not get useful results due to the low lipid yield for meaningful interpretation.

More homogeneous shape without changes in raw materials reflects the clay source might be the same but a preferred shape was emphasized during the Chinese contact period.

The result of shape analysis suggests that there might be the emergence of specialized groups after the contact with the Europeans. This supports our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society. This inequality might develop based on the monopolies for trade between a small number of indigenous people and the social organization switch from a more corporate to a more network organization [@Feinman2000].

The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery among those contact phases. Compared to other region in Taiwam, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

>>>>

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, particularly trade ornaments, during the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization as an indicator of the presence of ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. There are significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape based on the petrographic analysis. More standardized shape in the Chinese period might indicate the appearance of specialized group. The pottery shape during the Chinese period tend to be short with oval-shaped body showing wider body diameter compared to previous two phases. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].

Although there is differences in pottery shape between pre-European and post-European period, it is more significant in the Chinese period with more standardized pottery shape. The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society, however, stronger in the Chinese period instead of the European period that we originally thought to be. The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].
